---
title: "Practical exercises: solutions"
author: "Tobias Rüttenauer"
date: "July 02, 2022"
output_dir: docs
output: 
  html_document:
    theme: flatly
    highlight: haddock
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
theme: united
bibliography: sicss-spatial.bib
link-citations: yes
---

### Required packages

```{r, message = FALSE, warning = FALSE, results = 'hide'}
pkgs <- c("sf", "nngeo", "mapview", "nomisr", "tmap", "tmaptools", 
          "osmdata", "randomForest") # note: load spdep first, then spatialreg
lapply(pkgs, require, character.only = TRUE)

```

### Session info

```{r}
sessionInfo()

```
  
### Load spatial data

See previous file.

```{r}
load("_data/msoa_spatial.RData")

```

# Exercise 1

Add a variable for the distance to Boris Johnson's home (10 Downing St, London SW1A 2AB, UK). 

    * You could either find the coordinates manually or you use `tmaptools` function `geocode_OSM()` using OpenStreetMaps. 
    
    * There are also Google Maps APIs like `ggmap` or `mapsapi` but they require registration.
    
    * You can also search for something completely different.

```{r}
# Geocode an address using tmaptools
adr <- "10 Downing St, London SW1A 2AB, UK"
boris.spdf <- geocode_OSM(adr, as.sf = TRUE)
mapview(boris.spdf)

# Transform into same projection
boris.spdf <- st_transform(boris.spdf, crs = st_crs(msoa.spdf))

# Copute distances betweens msoas and Boris 
msoa.spdf$dist_boris <- as.numeric(st_distance(msoa.spdf, boris.spdf))

```

# Exercise 2

Please think about other potential variables you could add to the previous data which might be relevant for predicting NO2.

  * You can for instance use the census API, or the OpenStreetMap API
  
  * You can also search for other data in the [London Datastore](https://data.london.gov.uk/dataset/cultural-infrastructure-map) 
  
### Census API  

```{r}
### Add census data, country of origin (KS204EW)

london_ids <- msoa.spdf$MSOA11CD
x <- nomis_data_info()

# Get internal ids
stats <- c("KS204EW")
oo <- which(grepl(paste(stats, collapse = "|"), x$name.value))
ksids <- x$id[oo]
ksids # This are the internal ids


# Query data in loop over the required statistics
for(i in ksids){

  # Determin if data is divided by sex or urban-rural
  nd <- nomis_get_metadata(id = i)
  if("RURAL_URBAN" %in% nd$conceptref){
    UR <- TRUE
  }else{
    UR <- FALSE
  }
  if("C_SEX" %in% nd$conceptref){
    SEX <- TRUE
  }else{
    SEX <- FALSE
  }

  # make data request
  if(UR == TRUE){
    if(SEX == TRUE){
      tmp_en <- nomis_get_data(id = i, time = "2011", 
                               geography = london_ids, # replace with "TYPE297" for all MSOAs
                               measures = 20100, RURAL_URBAN = 0, C_SEX = 0)
    }else{
      tmp_en <- nomis_get_data(id = i, time = "2011", 
                               geography = london_ids, # replace with "TYPE297" for all MSOAs
                               measures = 20100, RURAL_URBAN = 0)
    }
  }else{
    if(SEX == TRUE){
      tmp_en <- nomis_get_data(id = i, time = "2011", 
                               geography = london_ids, # replace with "TYPE297" for all MSOAs
                               measures = 20100, C_SEX = 0)
    }else{
      tmp_en <- nomis_get_data(id = i, time = "2011", 
                               geography = london_ids, # replace with "TYPE297" for all MSOAs
                               measures = 20100)
    }

  }

  # Append (in case of different regions)
  ks_tmp <- tmp_en

  # Make lower case names
  names(ks_tmp) <- tolower(names(ks_tmp))
  names(ks_tmp)[names(ks_tmp) == "geography_code"] <- "msoa11"
  names(ks_tmp)[names(ks_tmp) == "geography_name"] <- "name"

  # replace weird cell codes
  onlynum <- which(grepl("^[[:digit:]]+$", ks_tmp$cell_code))
  if(length(onlynum) != 0){
    code <- substr(ks_tmp$cell_code[-onlynum][1], 1, 7)
    if(is.na(code)){
      code <- i
    }
    ks_tmp$cell_code[onlynum] <- paste0(code, "_", ks_tmp$cell_code[onlynum])
  }

  # save codebook
  ks_cb <- unique(ks_tmp[, c("date", "cell_type", "cell", "cell_code", "cell_name")])

  ### Reshape
  ks_res <- tidyr::pivot_wider(ks_tmp, id_cols = c("msoa11", "name"),
                               names_from = "cell_code",
                               values_from = "obs_value")

  ### Merge
  if(i == ksids[1]){
    census_keystat.df <- ks_res
    census_keystat_cb.df <- ks_cb
  }else{
    census_keystat.df <- merge(census_keystat.df, ks_res, by = c("msoa11", "name"), all = TRUE)
    census_keystat_cb.df <- rbind(census_keystat_cb.df, ks_cb)
  }

}


# Merge with MSOA geometries above
msoa.spdf <- merge(msoa.spdf, census_keystat.df, 
                   by.x = "MSOA11CD", by.y = "msoa11", all.x = TRUE)

# Percent non-EU residents
msoa.spdf$per_eu <- (msoa.spdf$KS204EW_100 + msoa.spdf$KS204EW_200) / msoa.spdf$KS204EW0001 * 100
msoa.spdf$per_non_eu <- 100 - msoa.spdf$per_eu

```



### OpenStreetMaps  

```{r}

# Bounding box
q <- opq(bbox = st_bbox(st_transform(msoa.spdf, 4326)))

# Lets get industrial landuse areas
osmq <- add_osm_feature(q, key = "highway", value = c("motorway", "trunk"))

# Query
highway.osm <- osmdata_sf(osmq)

# Make unique points / polygons 
highway.osm <- unique_osmdata(highway.osm)

# Get the lines is this case
highway.spdf <- highway.osm$osm_lines

# Drop OSM file
rm(highway.osm); gc()

plot(st_geometry(highway.spdf))


# Calculate distance for each MSOA
# Use geometric centroid of each MSOA
cent.sp <- st_centroid(msoa.spdf[, "MSOA11CD"])

# get into sam projection
highway.spdf <- st_transform(highway.spdf, st_crs(msoa.spdf))

# Get K nearest neighbour with distance
knb.dist <- st_nn(cent.sp, highway.spdf,
                  k = 1, returnDist = TRUE,
                  progress = FALSE)
msoa.spdf$dist_highway <- unlist(knb.dist$dist)

mapview(msoa.spdf[, "dist_highway"])
```


### London Data store

```{r}
# Get low emission neighbourhood
lowemission.link <- "https://data.london.gov.uk/download/low_emission_neighbourhoods/a1eb1018-8933-4e69-96ca-97402637ad92/Low_Emission_Neighbourhoods.gpkg"

# This time, we have Geopackage format (gpkg)
tmpf <- tempfile(fileext = ".gpkg")
download.file(lowemission.link, destfile = tmpf, mode = "wb")

# And we only load the layer containing pubs
st_layers(tmpf)
lowemission.spdf <- st_read(dsn = tmpf)
unlink(tmpf)

# Someting is wrong here with projection! Assign manually
st_crs(lowemission.spdf) <- 27700

mapview(st_geometry(lowemission.spdf))

# MSOA intersects with low emission zone?
lowemission.spdf <- st_transform(lowemission.spdf, st_crs(msoa.spdf))
msoa.spdf$lez <- 0
within <- msoa.spdf[lowemission.spdf,]
msoa.spdf$lez[which(msoa.spdf$MSOA11CD %in% within$MSOA11CD)] <- 1
table(msoa.spdf$lez)

```

# Exercise 3

Given you have added new data, can you use some model to predict the importance of single variables for predicting NO2 levels across London?

```{r}
# Train a random forest
rf.mod <- randomForest(no2 ~ per_mixed + per_asian + per_black + per_other + 
                         per_owner + per_social + pubs_count + POPDEN + ulez + lez +
                         per_non_eu + dist_highway + dist_boris, 
                       data = st_drop_geometry(msoa.spdf), 
                       mtry = 2, 
                       ntree = 1000,
                       importance = TRUE)

# Inspect the mechanics of the model
varImpPlot(rf.mod)

# Partial dependence plots
imp <- importance(rf.mod)
impvar <- rownames(imp)[order(imp[, 1], decreasing=TRUE)]
op <- par(mfrow=c(2, 3))
for (i in seq_along(impvar)) {
    partialPlot(rf.mod, st_drop_geometry(msoa.spdf), impvar[i], xlab=impvar[i],
                main=paste("Partial Dependence on", impvar[i]),
                ylim=c(27, 34)
                )
}
par(op)
```


# Exercise 4

In case you still have some time left, can you draw a map of the spatial distribution of deprivation in London?

  * The [Index of Multiple Deprivation](https://www.gov.uk/guidance/english-indices-of-deprivation-2019-mapping-resources) offers several resources for mapping.
  
  * [Geoportal](https://geoportal.statistics.gov.uk/) is a great resource for geographical boundaries

```{r}
dn <- "_data"

# Read index of multiple deprivation
imd.link <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/845345/File_7_-_All_IoD2019_Scores__Ranks__Deciles_and_Population_Denominators_3.csv"
download.file(imd.link, paste0(dn, "/imd.csv"))
imd.df <- read.csv(paste0(dn, "/imd.csv"))

# Clean names
header <- make.names(tolower(names(imd.df)))
header <- gsub(".*\\.\\.imd\\.\\.(.*)$", "imd.\\1", header)
header <- gsub(".*\\.\\.idaci\\.\\.(.*)$", "idaci.\\1", header)
header <- gsub(".*\\.\\.idaopi\\.\\.(.*)$", "idaopi.\\1", header)
header <- gsub("education\\.\\.", "education\\.", header)
header <- gsub("\\.\\..*", "", header)
header <- gsub("rank\\.of\\.(.*)\\.score", "\\1.rank", header)
header <- gsub("rank\\.of\\.(.*)", "\\1.rank", header)
header <- gsub("crime\\.and\\.disorder", "crime", header)
header <- gsub("idaci", "income\\.children", header)
header <- gsub("idaopi", "income\\.older", header)
header <- gsub("children\\.and\\.young", "children\\.young", header)
header <- gsub("lsoa\\.code", "lsoa", header)
names(imd.df) <- header

### For now, we plot only London
# If you want to map for England, get the shape file here: https://geoportal.statistics.gov.uk/search?collection=Dataset&sort=name&tags=all(BDY_LSOA%2CDEC_2011)

# Read LSOA layer of shapefile
lsoa.spdf <- st_read(dsn = paste0(dn, "/statistical-gis-boundaries-london/ESRI"),
                     layer = "LSOA_2011_London_gen_MHW" 
                     )

# Merge London lsoa and imd
lsoa.spdf <- merge(lsoa.spdf, imd.df,
                   by.x = "LSOA11CD", by.y = "lsoa", all.y = FALSE)


# Map the index of multiple deprivation score
# Here we use a continous scale with tmap
mp3 <- tm_shape(lsoa.spdf) + 
  tm_fill(col = "imd.score", 
          style = "cont", # algorithm to def cut points
          palette = "inferno", # colours
          alpha = 1, # transparency 
          title = "Score", 
          legend.hist = FALSE # histogram next to map?
          ) +
  tm_layout(frame = FALSE,
            legend.frame = TRUE, legend.bg.color = TRUE,
            legend.position = c("right", "bottom"),
            legend.outside = FALSE,
            main.title = "Multiple Deprivation", 
            main.title.position = "center",
            main.title.size = 1.4,
            legend.title.size = 0.8,
            legend.text.size = 0.8)

mp3
```

