---
title: "Part 1: Using and linking spatial data"
author: "Tobias Rüttenauer"
date: "July 02, 2022"
output_dir: docs
output: 
  html_document:
    theme: flatly
    highlight: haddock
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
theme: united
bibliography: sicss-spatial.bib
link-citations: yes
---

### Required packages

```{r, message = FALSE, warning = FALSE, results = 'hide'}
pkgs <- c("sf", "mapview", "nngeo", "tmap", "tmaptools", "viridisLite") 
lapply(pkgs, require, character.only = TRUE)

```

### Session info

```{r}
sessionInfo()

```

### Load previous data

```{r}
load(file = "_data/msoa_spatial.RData")
```

A large advantage of spatial data is that different data sources can be connected and combined. Another nice advatange is: you can create very nice maps. And it's quite easy to do! Ilya Kashnitsky's [Dataviz course](https://github.com/ikashnitsky/dataviz-mpidr) gives a much more comprehensive introduction to maps.

Many packages and functions can be used to plot maps of spatial data. For instance, ggplot as a function to plot spatial data using `geom_sf()`. I am perconally a fan of tmap, which makes many steps easier (but sometimes is less flexible).

# Tmaps


For instance, lets plot the NO2 estimates using tmap.

```{r}

# Define colours
cols <- viridis(n = 7, direction = 1, option = "C")

mp1 <-  tm_shape(msoa.spdf) + 
  tm_fill(col = "no2", 
          style = "fisher", # algorithm to def cut points
          n = 7, # Number of requested cut points
          palette = cols, # colours
          alpha = 1, # transparency 
          title = "NO2", 
          legend.hist = FALSE # histogram next to map?
          ) +
  tm_borders(col = "white", lwd = 0.5, alpha = 0.5) 

mp1

```

Tmap allows to easily combine different objects by defining a new object via `tm_shape()`.

```{r}

# Define colours
cols <- viridis(n = 7, direction = 1, option = "C")

mp1 <-  tm_shape(msoa.spdf) + 
  tm_fill(col = "no2", 
          style = "fisher", # algorithm to def cut points
          n = 7, # Number of requested cut points
          palette = cols, # colours
          alpha = 1, # transparency 
          title = "NO2", 
          legend.hist = FALSE # histogram next to map?
          ) +
  tm_borders(col = "white", lwd = 0.5, alpha = 0.5) +
  tm_shape(ulez.spdf) +
  tm_borders(col = "red", lwd = 1, alpha = 1) 

mp1

```

And it is easy to change the layout.

```{r}

# Define colours
cols <- viridis(n = 7, direction = 1, option = "C")

mp1 <-  tm_shape(msoa.spdf) + 
  tm_fill(col = "no2", 
          style = "fisher", # algorithm to def cut points
          n = 7, # Number of requested cut points
          palette = cols, # colours
          alpha = 1, # transparency 
          title = expression('in'~mu*'g'/m^{3}), 
          legend.hist = FALSE # histogram next to map?
          ) +
  tm_borders(col = "white", lwd = 0.5, alpha = 0.5) +
  tm_shape(ulez.spdf) +
  tm_borders(col = "red", lwd = 1, alpha = 1) +
  tm_layout(frame = FALSE,
            legend.frame = TRUE, legend.bg.color = TRUE,
            legend.position = c("right", "bottom"),
            legend.outside = FALSE,
            main.title = "NO2", 
            main.title.position = "center",
            main.title.size = 1.6,
            legend.title.size = 0.8,
            legend.text.size = 0.8)

mp1

```

Obviously, we can also add some map information from OSM. However, it's sometime a bit tricky with the projection. That's why we switch into the OSM projection here.

```{r}
# Save old projection
crs_orig <- st_crs(msoa.spdf)

# Change projection
ulez.spdf <- st_transform(ulez.spdf, 4326)
msoa.spdf <- st_transform(msoa.spdf, 4326)

# Get OSM data for background
osm_tmp <- read_osm(st_bbox(msoa.spdf), ext = 1, 
                    type = "stamen-toner", zoom = NULL) 

# Define colours
cols <- viridis(n = 7, direction = 1, option = "C")

mp1 <-  tm_shape(osm_tmp) + tm_rgb() +
  tm_shape(msoa.spdf) + 
  tm_fill(col = "no2", 
          style = "fisher", # algorithm to def cut points
          n = 7, # Number of requested cut points
          palette = cols, # colours
          alpha = 0.8, # transparency 
          title = expression('in'~mu*'g'/m^{3}), 
          legend.hist = FALSE # histogram next to map?
          ) +
  #tm_borders(col = "white", lwd = 0.5, alpha = 0.5) +
  tm_shape(ulez.spdf) +
  tm_borders(col = "red", lwd = 1, alpha = 1) +
  tm_layout(frame = FALSE,
            legend.frame = TRUE, legend.bg.color = TRUE,
            legend.position = c("right", "bottom"),
            legend.outside = FALSE,
            main.title = "NO2", 
            main.title.position = "center",
            main.title.size = 1.6,
            legend.title.size = 0.8,
            legend.text.size = 0.8)

mp1

```

Tmap also makes it easy to combine single maps 

```{r}
# Define colours
cols1 <- viridis(n = 7, direction = 1, option = "C")

# Define colours
cols2 <- viridis(n = 7, direction = 1, option = "D")

mp1 <-  tm_shape(osm_tmp) + tm_rgb() +
  tm_shape(msoa.spdf) + 
  tm_fill(col = "no2", 
          style = "fisher", # algorithm to def cut points
          n = 7, # Number of requested cut points
          palette = cols1, # colours
          alpha = 0.8, # transparency 
          title = expression('in'~mu*'g'/m^{3}), 
          legend.hist = FALSE # histogram next to map?
          ) +
  #tm_borders(col = "white", lwd = 0.5, alpha = 0.5) +
  tm_shape(ulez.spdf) +
  tm_borders(col = "red", lwd = 1, alpha = 1) +
  tm_layout(frame = FALSE,
            legend.frame = TRUE, legend.bg.color = TRUE,
            legend.position = c("right", "bottom"),
            legend.outside = FALSE,
            main.title = "NO2", 
            main.title.position = "center",
            main.title.size = 1.4,
            legend.title.size = 0.8,
            legend.text.size = 0.8)

mp2 <-  tm_shape(osm_tmp) + tm_rgb() +
  tm_shape(msoa.spdf) + 
  tm_fill(col = "per_black", 
          style = "fisher", # algorithm to def cut points
          n = 7, # Number of requested cut points
          palette = cols2, # colours
          alpha = 0.8, # transparency 
          title = "% white", 
          legend.hist = FALSE # histogram next to map?
          ) +
  #tm_borders(col = "white", lwd = 0.5, alpha = 0.5) +
  tm_shape(ulez.spdf) +
  tm_borders(col = "red", lwd = 1, alpha = 1) +
  tm_layout(frame = FALSE,
            legend.frame = TRUE, legend.bg.color = TRUE,
            legend.position = c("right", "bottom"),
            legend.outside = FALSE,
            main.title = "Ethnic Black inhabitants", 
            main.title.position = "center",
            main.title.size = 1.4,
            legend.title.size = 0.8,
            legend.text.size = 0.8)

tmap_arrange(mp1, mp2, ncol = 2, nrow = 1)
```


# References